\Large \bf DRAFT: Direct-style process creation on Linux

### 2021-04-07 research/paper/paper.tex
paper

\usepackage{tikz}
\usepackage{amsmath}

@@

@@

\section{Introduction}\label{introduction}
Existing process creation interfaces can be divided into three categories:
Fork-style, spawn-style, and direct-style.

Fork-style (such as \texttt{fork}) and spawn-style (such as \texttt{posix\_spawn}) are the most popular,
but have a number of disadvantages.
Correct, performant use of fork-style places significant constraints on the calling process,
and also has a complex programming model.
Spawn-style lifts many of these constraints, and is simpler to program against,
but is fundamentally limited in the kinds of processes it can create.
With both fork-style and spawn-style,
it is difficult to
branch based on results or errors encountered while initializing the new process:
fork-style requires IPC if it wishes to communicate errors back to the rest of the program,
and robust indication of errors in spawn-style requires substantial growth in the spawn interface.

Direct-style is more rarely seen, but has a number of advantages.
Direct-style involves creating a new child process in a non-running state,
and then calling individual system calls from the parent to initialize the child.
Direct-style does not constrain the calling process,
easily branches at any step of the initialization,
and can create any kind of process supported by the underlying system.

Our contribution is an implementation of direct-style process creation for Linux,
making use of our work on the \texttt{rsyscall} project.
The \texttt{rsyscall} project develops language-specific ``process-independent'' system call libraries for Linux,
which replace libc system call wrappers, and allow a program to perform cross-process system calls.
rsyscall currently has best support for Python,
but can be ported to other languages;
the examples we show in this paper will be in Python,
but generalize easily to other languages.

Our direct-style process creation implementation uses \texttt{clone} to create a child process in an embryonic state
which can then be manipulated from user code using syscalls with \texttt{rsyscall}.
A direct-style \texttt{clone} has all the advantages of direct-style process creation
over Linux's fork-style \texttt{fork} and spawn-style \texttt{posix\_spawn}:
\begin{compactitem}
\item
  Processes can be created correctly and performantly from any parent process,
  with a straightforward, familiar API.
\item
  Processes can be created with the full set of Linux features.
\item
  Syscall results are reported individually to the parent process,
  rather than batched together as with spawn-style or manually reported via IPC as with fork-style.
\end{compactitem}

We've implemented a variety of useful applications using direct-style process creation on Linux,
which we'll demonstrate in this paper.
Direct-style \texttt{clone} has minimal performance cost in normal situations,
and can outperform \texttt{fork} in some cases.
We have deployed direct-style process creation in production internally at \twosigma.

Our implementation is distributed as a feature of \texttt{rsyscall},
and is immediately deployable on recent Linux kernels.
\texttt{rsyscall} is open source and available from
[Github url].
% \url{https://github.com/catern/rsyscall}.

This paper is organized in several sections.
\begin{compactitem}
\item In section \ref{background}, we provide background on fork-style, spawn-style, and direct-style process creation,
and consider their advantages and disadvantages in more depth.
\item In section \ref{overview}, we discuss the design of the API,
including our usage of \texttt{rsyscall}.
\item In section \ref{applications}, we show examples of usage of direct-style process creation in Python on Linux,
including use of process features which are difficult to use on Linux with \texttt{fork} or \texttt{posix\_spawn}.
\item In section \ref{implementation}, we explore our implementation in depth,
and discuss several difficult process-creation related details of \texttt{rsyscall}'s implementation.
\item In section \ref{evaluation}, we evaluate, somehow (TODO).
\item In section \ref{related_work},
we examine other work which also touches on the topics covered by this paper.
\item In section \ref{future_work},
we list several directions for future work related to direct-style process creation and to \texttt{rsyscall}.
\item In section \ref{conclusions}, we briefly outline our conclusions about direct-style and the future of process creation.
\end{compactitem}
\section{Background}\label{background}
As mentioned in the introduction,
on most operating systems today,
the available interfaces for process creation
are either "fork-style" or "spawn-style".
A few operating systems have a third style, which we refer to as "direct-style".
\subsection{Fork-style}
By far the most well-known fork-style interface is Unix's fork.\cite{forkhist}
With a fork-style interface,
a process, called the parent process, calls \texttt{fork},
and a copy of that process is made,
called the child process,
differing only in a few attributes such as process id.
The same program continues executing in both processes,
with a different return value from fork to allow it to distinguish whether it is running in the parent or the child.
Typically, the program will case on the return value
and, if running in the child,
call various system calls to modify the state of the new process.

Fork has a number of flaws,
and much ink has been spilled detailing them.
We'll cover the most prominent briefly here,
but see Baumann 2019 \cite{forkroad} for a more detailed evaluation.

@@

However, we believe that a natural port of direct-style process creation to Linux
provides an interface that is just as simple to use as fork-style or spawn-style on Linux,
without their disadvantages.
Such an interface, like fork-style and spawn-style,
would implicitly copy some attributes to the new process,
rather than being capability-oriented and requiring all attributes to be specified explicitly.
This is the kind of direct-style interface that we contribute in this paper.
\section{Overview}\label{overview}
In most operating system APIs, including the ones available on Linux,
when a program makes a syscall, it implicitly operates on the current process.
To perform direct-style process creation on Linux,
we need an API where we instead explicitly specify in which process we want to make a syscall.

\texttt{rsyscall} provides this API.
The \texttt{rsyscall} project develops
language-specific, object-capability-model, low-abstraction libraries for the Linux syscall interface,
bypassing libc.

To perform direct-style process creation,
we use \texttt{rsyscall} to call \texttt{clone} on the ``local'' process to create a new child process.
We call various syscalls in the child process to mutate it until it reaches the desired state,
at which point we can call \texttt{execve} in the child process to start it running on its own.
After that, the process is no longer under our control, and it functions like any other child process.
The child process can be monitored using normal Linux child monitoring syscalls,
such as \texttt{waitid}.

\lstinputlisting[float=*ht,language=Python,label={basic},caption={Creating a new process, changing CWD, and execing}]
{examples/basic/direct.py}
Listing \ref{basic} shows this in action in Python.
We create a new child process using \texttt{clone} on the local process \verb|local|,
and receive back a capability to control it.
We call \texttt{chdir} in the child process to change its working directory to \texttt{/dev}.
Then we call \texttt{execve},
passing the path of the \texttt{cat} executable,
the arguments to the new executable,
and the unchanged set of environment variables.
The name of the executable is passed as the first argument,
as is typical in Unix.
Calling \texttt{execve} consumes the capability,
and later use of this child process capability will fail with an exception.

The behavior of system calls made on a remote process through \texttt{rsyscall}
is identical to the behavior of system calls made implicitly on the current process through libc.
This is an API that systems programmers are already familiar with,
and which is already thoroughly documented through Linux system call manpages.
The novelty is constrained to the newly explicit process capabilities.

The lack of novelty in the API is important not only for the learning curve,
but also for power.
Since we are using conventional Linux system calls,
just in a cross-process way,
direct-style \texttt{clone} will always be at least as powerful as \texttt{fork}.
\section{Applications}\label{applications}
In this section,
we'll discuss several more advanced uses of processes
using more sophisticated Linux features,
and accompany them with Python examples using direct-style process creation.
All of these examples use only conventional Linux system calls,
and can therefore translated straightforwardly into any language with an \texttt{rsyscall} API.
For concision, we assume that we are running with sufficient permissions in these examples,
but these examples can all be run without privileges with appropriate use of \texttt{CLONE.NEWUSER}.

Implementing these examples with \texttt{fork} or \verb|posix_spawn| requires substantially more boilerplate,
or even a complete change in implementation approach,
such as in section \ref{shared_fd_table}.
We'll compare our direct-style implementations in this section
to fork-style implementations
in section \ref{evaluation}.

Several of these examples demonstrate functionality that can be done today by specialized software and services.
Such software often allows use of one or two of these techniques, but prevents the use of others.
For example, a shell allows the creation of pipelines and container systems like Docker allow sandboxing,
but the two are difficult to combine.

One of the key advantages of direct-style process creation is that the techniques we demonstrate here
can be easily used in combination,
to meet the specific needs of individual real-world applications.
\subsection{Abstraction through file descriptor passing}\label{fd_abstraction}
In Linux (and Unix in general),
a program can pass file descriptors from the parent process to a child process
by using file descriptor inheritance.
All open file descriptors are copied at process creation time from the parent process to the child process;
the child process can then make independent use of the file descriptors.
We'll discuss this important feature further in section \ref{fdtables}.

Most resources in Linux are managed through file descriptors,
so this feature allows the parent process to pass a wide variety of resources to the child process.
An executable program can be provided a file descriptor,
which can correspond to any file in any filesystem,
or to a network connection\cite{ucspi},
or a pipe connected to another process,
or a variety of other resources,
without the process being aware of what specific resource it is accessing.
In this way, we immediately know "for free"\cite{theoremsforfree}
that the behavior of the program does not depend on the name of the file,
or the hostname to it which the socket is connected,
or the executable running in the process on the other end of the pipe,
or other abstracted-away details.

\lstinputlisting[float=*ht,language=Python,label={fds},caption={Passing down FDs}]
{examples/fds/direct.py}
In Listing \ref{fds},
we first open a file with read-write permission.
Then we create a child process,
which inherits all file descriptors from its parent.
We indicate that we want to operate on the child's inherited copy of the file descriptor with \verb|child.inherit_fd|.
Then we disable \texttt{CLOEXEC} on the child's file descriptor,
which on Linux we can do by clearing the FD flags with \verb|fcntl(fd, F_SETFD, 0)|.
Finally, we execute some executable,
passing the file descriptor number as an argument;
since we disabled \texttt{CLOEXEC} on the file descriptor,
the executable will be able to use that file descriptor through that file descriptor number.
Unlike in listing \ref{basic}, we call the \texttt{execv} wrapper rather than \texttt{execve},
so that we don't need to pass the (unmodified) environment to every \texttt{exec} call in these examples.
\subsection{Non-shared-memory concurrency}
Processes run concurrently,
which allows exploiting the parallelism of the underlying hardware.
Since processes don't share memory,
they can provide a less complex parallel programming environment
than shared-memory thread-based approaches.

@@

File descriptor passing over a Unix domain socket
can allow similar behavior without sharing the file descriptor table,
but is substantially more complex for the user.
Nevertheless, for fork-style and spawn-style process creation,
file descriptor passing over a Unix domain socket
is our only option to do something like this.
\section{Implementation}\label{implementation}
\subsection{Basics about \texttt{rsyscall}}
In this section, we'll give a brief overview of the implementation of \texttt{rsyscall},
and then focus on implementation issues specific to process creation.

\texttt{rsyscall} can be conceptually divided in two parts:
the basic cross-process syscall primitive,
and a language-specific library built on top.
The Python language-specific library has already been demonstrated above.
Such libraries only need to be able to call syscalls and explicitly specify a process in some way;
they are, for the most part, agnostic to how the cross-process syscall is implemented.

On Linux \verb|x86_64|, a syscall is specified by a syscall number plus six register-sized arguments;
a syscall returns one register-sized value.
\texttt{rsyscall}'s default implementation of cross-process syscalls sends those seven integers over a pipe,
and waits for a response on another pipe.
Processes are created running a loop which, at each iteration,
reads a syscall request off the pipe,
performs that syscall,
and writes the return value back over the return pipe.
In this way, a cross-process syscall works much like a very primitive remote procedure call.

We use pipes (or other file descriptors) for transport of data rather than a shared memory segment
to support straightforward integration with event loops.

ptrace provides an alternative means to perform arbitrary actions on other processes,
including system calls.
However, among other issues, it has the substantial disadvantage of not permitting multiple ptracers.
A ptrace-based implementation would prevent using \texttt{strace} or \texttt{gdb} on \texttt{rsyscall}-controlled processes,
which is an unacceptable limitation for a general-purpose utility.

@@

The most simple way to access memory is for the local process to be in the same address space as the target process.
This is the case most of the time; we pass \texttt{CLONE.VM} to \texttt{clone} by default in the Python wrapper.
However, the target process may be in another address.
For example, if the target process is at a different privilege level,
it must be in a different address space,
as sharing an address space with a process at a different privilege level is unsafe.

We implement access to the memory of processes in different address spaces through another set of pipes,
by copying memory into and out of those pipes using \texttt{read} and \texttt{write} system calls in the target process.
A \texttt{read} syscall corresponds to a write to memory,
and a \texttt{write} syscall corresponds to a read from memory.

The \verb|process_vm_readv| and \verb|process_vm_writev| system calls
allow the caller to read and write memory from the virtual address space of other processes.
However, they require that the caller have specific credentials relative to the process being accessed,
which may not always be the case.
Additionally, these system calls are disabled if ptrace is disabled system-wide,
which is a niche but possible system configuration.
To ensure that \texttt{rsyscall} can be used for arbitrary purposes and on arbitrary systems, we avoided these calls.
\subsection{\texttt{clone}}\label{clone}
Now that we've established the basic implementation details of \texttt{rsyscall},
we'll consider specific issues related to process creation and initialization.

There are three Linux system calls which create processes:
\texttt{fork}, \texttt{vfork} and \texttt{clone}.
\texttt{vfork} suspends execution of the parent process
while waiting for the child process to exit or call \texttt{execve},
which is immediately unsuitable for direct-style process creation.
\texttt{clone} provides a superset of the functionality of \texttt{fork},
so we focused our attention on \texttt{clone}.

\texttt{clone}
(along with \texttt{fork}) creates a new process
which immediately starts executing at the next instruction after the syscall instruction,
in parallel with the parent process,
with its registers in generally the same state as the parent process.

Common Linux libcs provide a wrapper for \texttt{clone}
which allow executing an arbitrary function with an arbitrary argument,
but since we want to support remote execution,
we can't use the libc wrapper.

Instead, since \texttt{clone} lets us change the stack for the new process,
we use this to return, after the system call to an arbitrary function address stored on the stack.
Further arguments can also be passed on the stack,
with the aid of a trampoline to match C calling conventions if necessary.

We use this to create processes running the syscall request-response loop described in the previous section.
After this,
most system calls can be called as normal.
The new process can be modified freely through \texttt{chdir}, \texttt{dup2}, and other system calls.
Out of system calls related to process creation,
only \texttt{execve} and \texttt{unshare} need substantial further attention.
\subsection{\texttt{exec}}\label{execve}
\texttt{execve} is unusual and requires careful design,
because when it is successful, it does not return.
Therefore we need a way to determine if \texttt{execve} is successful;
naively waiting for a response to the syscall request will leave us waiting forever.

There is a traditional technique to detect a successful \texttt{execve} using a pipe.
We create a pipe before forking,
ensure both ends are marked \verb|O_CLOEXEC|,
perform the fork,
call \texttt{execve} in the child,
close the write end of the pipe,
and wait for EOF on the read end.
If the child process has neither successfully called \texttt{execve}, nor exited for some other reason,
then the write end of the pipe will still be open in the child process's fd table,
and the read end of the pipe will not return EOF.
But once the child process calls \texttt{execve} successfully,
\verb|O_CLOEXEC| will cause the write end of the pipe to be closed,
and the read end of the pipe will return EOF.

This trick works well with \texttt{fork};
but it's not general enough to work with \texttt{clone}.
Child processes can be created with the \verb|CLONE_FILES| flag passed to \texttt{clone},
which causes the parent process and child process to share a single fd table;
we showed an example of this in section \ref{shared_fd_table}.
This means that when the parent process closes the write end of the pipe,
it will also be closed in the child process,
and the read end of the pipe will immediately return EOF,
regardless of whether the child has called \texttt{execve} or exited.

Fortunately, there is an alternative solution, which does work with \verb|CLONE_FILES|.
The \texttt{ctid} argument to \texttt{clone} specifies a memory address which,
when the \verb|CLONE_CHILD_CLEARTID| flag is set,
the kernel sets to zero when the child exits or execs,
and then, crucially, performs a futex wakeup on.

Futexes are usually used for the implementation of userspace shared-memory synchronization constructs,
but the relevant detail for us here is that we can wait on an address
until a futex wakeup is performed on that address.
This means we can wait on \texttt{ctid} until the futex wakeup is performed,
and in this way get notified of the child process calling \texttt{execve} or exiting.

Unfortunately, futexes in current Linux integrate poorly:
There is no way for a single process to wait for more than one futex at a time,
and no way to monitor a futex with file-descriptor-monitoring syscalls such as \texttt{poll}.
The best we can do is create a dedicated child process for each futex we want to wait on,
and have this child process exit when the futex has a wakeup.
Monitoring child processes can be straightforwardly integrated into an event loop.

While slightly complex to implement, this solution works well.
We provide \texttt{ctid} whenever we call \texttt{clone},
and set up a process to wait on that futex.
Then, when we call any syscall,
we wait for either the syscall to return an error or the futex process to exit,
whichever comes first.
Either will normally indicate that the syscall fails,
but if the futex process exits,
without the child process itself exiting,
then we can catch that in \texttt{execve} and know that the child has successfully called \texttt{execve}.

If the futex process and child process both exit,
it's ambiguous whether the child process successfully called \texttt{execve};
this ambiguity is unfortunate, but it is also present in the pipe-based approach.
This is, we believe, the best solution currently available.

We would prefer for Linux to natively provide functionality to wait for a child's \texttt{execve}.
Some other Unix-like systems provide this;
kqueue, on FreeBSD, allows waiting for exec in arbitrary processes through kqueue's \verb|EVFILT_PROC|.

One approach for Linux would be to add a new \texttt{clone} flag to
opt-in to receiving \texttt{WEXECED} events through \texttt{waitid}.
A new \texttt{waitid} flag alone is not sufficient,
since, to integrate this feature into an event loop,
it's necessary to receive \texttt{SIGCHLD} signals or readability notifications on a pidfd
when the \texttt{WEXECED} event happens.
\subsection{Managing file descriptor tables}\label{fdtables}
As demonstrated in section \ref{shared_fd_table},
the \verb|CLONE_FILES| flag can be passed to \texttt{clone},
causing the file descriptor table to be shared between the parent process and child process.
The same file descriptors are open in both processes at the same numbers,
and if new file descriptors are opened in either process,
they are also visible in the other process.
The file descriptor table is unshared once the child calls \texttt{execve}.
This is simple to model and convenient for many purposes.

>>>>

\usepackage{tikz}
\usepackage{dblfloatfix}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{pifont}
\usepackage{comment}

\newcommand{\cmark}{\ding{51}}
\newcommand{\xmark}{\ding{55}}

@@

% Ugh, these aren't the same size right now.
\newcommand{\tbad}{\xmark:}
\newcommand{\tgood}{\cmark:}
% \newcommand{\tbad}{\text{\sffamily X} :}
% \newcommand{\tgood}{\checkmark:}

@@

\section{Introduction}\label{introduction}
Existing process creation interfaces can be divided into three categories:
Fork-style, spawn-style, and direct-style.
\begin{table*}
\begin{tabular}{l|l|l|l}
% BEGIN RECEIVE ORGTBL compare
 & fork & spawn & direct-style\\
\hline
Requirements on parent process & \tbad Single-threaded, small memory & \tgood None & \tgood None\\
Programming model & \tbad Complex (returns twice) & \tgood Simple (single call) & \tgood Simple (imperative)\\
Maximally powerful & \tgood Yes, can call any syscall & \tbad No, limited interface & \tgood Yes, can call any syscall\\
Reporting of results & \tbad Requires IPC & \tbad Not fine-grained & \tgood From individual child syscalls\\
Non-inherited attributes & Mutated by code in child & Set by arguments & Mutated by code in parent\\
% END RECEIVE ORGTBL compare
\end{tabular}
\caption{Features of fork-style vs spawn-style vs direct-style}
\label{tab:styles}
\end{table*}
\begin{comment}
#+ORGTBL: SEND compare orgtbl-to-latex :splice t
|                                | fork                                | spawn                       | direct-style                          |
|--------------------------------+-------------------------------------+-----------------------------+---------------------------------------|
| Requirements on parent process | \tbad Single-threaded, small memory | \tgood None                 | \tgood None                           |
| Programming model              | \tbad Complex (returns twice)       | \tgood Simple (single call) | \tgood Simple (imperative)            |
| Maximally powerful             | \tgood Yes, can call any syscall    | \tbad No, limited interface | \tgood Yes, can call any syscall      |
| Reporting of results           | \tbad Requires IPC                  | \tbad Not fine-grained      | \tgood From individual child syscalls |
| Non-inherited attributes       | Mutated by code in child            | Set by arguments            | Mutated by code in parent             |
% $
\end{comment}
Table \ref{tab:styles} summarizes the differences between each style.

Fork-style (such as \texttt{fork}) and spawn-style (such as \texttt{posix\_spawn}) are the most popular,
but have a number of disadvantages.
Correct, performant use of fork-style places significant constraints on the calling process,
and also has a complex programming model.
Spawn-style lifts many of these constraints, and is simpler to program against,
but is fundamentally limited in the kinds of processes it can create.
With both fork-style and spawn-style,
it is difficult to
branch based on results or errors encountered while initializing the new process:
fork-style requires IPC if it wishes to communicate errors back to the rest of the program,
and robust indication of errors in spawn-style requires substantial expansion of the spawn interface.

Direct-style is more rarely seen, but gives a number of advantages.
Direct-style involves creating a new child process in a non-running state,
and then calling individual system calls from the parent to initialize the child.
Direct-style does not constrain the calling process,
easily branches at any step of the initialization,
and can create any kind of process supported by the underlying system.

We'll discuss each style in more detail in section \ref{background}.

To perform direct-style process creation on Linux,
we need an API where we explicitly specify in which process we want to make a syscall.
This differs from the normal mode of operation on Linux,
where programs making syscalls implicitly operate on the current process.

Our contribution in this paper is this API, through the \texttt{rsyscall} library.
The \texttt{rsyscall} project develops
language-specific, object-capability-model, low-abstraction libraries for the Linux syscall interface,
bypassing libc.
\texttt{rsyscall} is currently focused on supporting Python,
but can be ported to other languages;
the examples we show in this paper will be in Python,
but generalize easily to other languages.
\texttt{rsyscall} is open source and available from
[Github url].

To perform direct-style process creation,
we use \texttt{rsyscall} to call \texttt{clone} on the ``local'' process to create a new child process.
We call various syscalls in the child process to mutate it until it reaches the desired state,
at which point we can call \texttt{exec} in the child process to start it running on its own.
After that, the process is no longer under our control, and it functions like any other child process.
The child process can be monitored using normal Linux child monitoring syscalls,
such as \texttt{waitid}.

These child process capabilities are created and managed in userspace,
by launching new processes running a syscall server,
which receives syscall requests, performs the syscall, and returns the response.
No kernel modifications are required, and \texttt{rsyscall} is immediately deployable on recent Linux kernel.
We discuss the implementation in depth in section \ref{implementation},
and examine several difficult process-creation related details.

\lstinputlisting[float=h,language=Python,label={basic},caption={Create new process, change CWD, and exec}]
{examples/basic/direct.py}
Listing \ref{basic} shows a program using \texttt{rsyscall}.
We create a new child process using \texttt{clone} on the local process \verb|local|,
and receive back a capability to control it.
We call \texttt{chdir} in the child process to change its working directory to \texttt{/dev}.
Then we call \texttt{execv} in the child,
passing the path of the \texttt{cat} executable
and the arguments to the new executable.
The name of the executable is passed as the first argument,
as is typical in Unix.
Calling \texttt{execv} consumes the capability,
and later use of this child process capability will fail with an exception.

The behavior of system calls made on a remote process through \texttt{rsyscall}
is identical to the behavior of system calls made implicitly on the current process through libc.
This is an API that systems programmers are already familiar with,
and which is already thoroughly documented through Linux system call manpages.
The novelty is constrained to the newly explicit process capabilities.
Besides flattening the learning curve,
this also ensures that direct-style \texttt{clone} will always be at least as powerful as \texttt{fork}.

We show more examples of direct-style process creation in section \ref{examples},
including use of process features which are difficult to use on Linux with \texttt{fork} or \texttt{posix\_spawn}.

Direct-style \texttt{clone} has acceptable performance cost,
and can outperform \texttt{fork} in some cases.
The performance overhead of creating processes with direct-style clone
is comparable to the performance cost of creating processes while inside a Linux namespace.
At \twosigma, we have deployed direct-style process creation in production
inside a library for distributed system deployment, Toplevel.
We evaluate in further detail in section \ref{evaluation}.
\section{Background on process-creation}\label{background}
As mentioned in the introduction,
on most operating systems today,
the available interfaces for process creation
are either "fork-style" or "spawn-style".
A few operating systems have a third style, which we refer to as "direct-style".

\subsection{Fork-style}
By far the most well-known fork-style interface is Unix's fork.\cite{forkhist}
With a fork-style interface,
a process, called the parent process, calls \texttt{fork},
and a copy of that process is made,
called the child process,
differing only in a few attributes such as process id.
The same program continues executing in both processes,
with a different return value from fork to allow it to distinguish whether it is running in the parent or the child.
Typically, the program will branch based on the return value
and, if running in the child,
call various system calls to modify the state of the new process.

Fork has a number of flaws,
and much ink has been spilled detailing them.
We'll cover the most egregious briefly here,
but see Baumann 2019 \cite{forkroad} for a more detailed evaluation.

@@

However, we believe that a natural port of direct-style process creation to Linux
provides an interface that is just as simple to use as fork-style or spawn-style on Linux,
without their disadvantages.
Such an interface, like fork-style and spawn-style,
would implicitly copy some attributes to the new process,
rather than being capability-oriented and requiring all attributes to be specified explicitly.
This is the kind of direct-style interface that we contribute in this paper.
\section{Examples}\label{examples}
In this section,
we'll demonstrate direct-style process creation
by demonstrating several advanced uses of processes
using more sophisticated Linux features.
For concision, we assume in these examples that we are running with sufficient permissions in these examples,
but these examples can all be run without privileges with appropriate use of \verb|CLONE_NEWUSER|.

These examples are not novel, as such, as they use only conventional Linux system calls;
but implementing them with \texttt{fork} or \verb|posix_spawn| requires substantially more code,
or even a complete change in implementation approach,
as described in section \ref{shared_fd_table}.
We'll compare our direct-style implementations in this section
to fork-style implementations
in section \ref{evaluation}.

Several of these examples provide functionality that can be done today by specialized software and services.
Such software often enables one or two of these uses of processes, but prevents others.
For example, a shell allows the creation of pipelines and container systems like Docker allow sandboxing,
but the two are difficult to combine.\cite{docker_pipe}

One of the key advantages of direct-style process creation is that the features we demonstrate here
can be easily used in combination,
to meet the specific needs of individual real-world applications.
\subsection{Abstraction through file descriptor passing}\label{fd_abstraction}
In Linux (and Unix in general),
a program can pass file descriptors from the parent process to a child process
by using file descriptor inheritance.
All open file descriptors are copied at process creation time from the parent process to the child process;
the child process can then make independent use of the file descriptors.
We'll discuss this important feature further in section \ref{fdtables}.

% TODO need to clarify this more, this is v2, Larry didn't like v1 beyond first sentence
Most resources in Linux are managed through file descriptors,
so this feature allows the parent process to pass a variety of resources to the child process,
such as files, network connections, pipes, or other resources.\cite{ucspi}
Since the resource is created outside the child process and passed down as an opqaue file descriptor,
the precise details and type of the resource are not available to the child.
This is a form of abstraction,
and we can know ``for free''\cite{theoremsforfree}
that the program running in the child process
will not rely on the details of how the file descriptor was created,
such as the filename that was opened or the hostname that was connected to.

\lstinputlisting[float=*ht,language=Python,label={fds},caption={Passing down FDs}]
{examples/fds/direct.py}
In Listing \ref{fds},
we first open a file with read-write permission in the local process.
Then we create a child process,
which inherits all file descriptors from its parent.
We indicate that we want to operate on the child's inherited copy of the file descriptor with \verb|child.inherit_fd|.
Then we disable \texttt{CLOEXEC} on the child's file descriptor,
which on Linux we can do by clearing the FD flags with \verb|fcntl(fd, F_SETFD, 0)|.
Finally, we execute some executable in the child process,
passing the file descriptor number as an argument;
since we disabled \texttt{CLOEXEC} on the file descriptor,
the executable will be able to use that file descriptor through that file descriptor number.
\subsection{Non-shared-memory concurrency}
Processes run concurrently,
which allows exploiting the parallelism of the underlying hardware.
Since processes don't share memory,
they can provide a less complex parallel programming environment
than shared-memory thread-based approaches.

@@

File descriptor passing over a Unix domain socket
can allow similar behavior without sharing the file descriptor table,
but is substantially more complex for the user.
Nevertheless, for fork-style and spawn-style process creation,
file descriptor passing over a Unix domain socket
is our only option.
\section{Implementation}\label{implementation}
In this section, we'll give a brief overview of the implementation of \texttt{rsyscall},
and then focus on implementation issues specific to process creation.

\texttt{rsyscall} can be conceptually divided in two parts:
the basic cross-process syscall primitive,
and a language-specific library built on top.
The Python language-specific library has already been demonstrated above.
Such libraries only need to be able to call syscalls and explicitly specify a process in some way;
they are, for the most part, agnostic to how the cross-process syscall is implemented.
The cross-process syscall primitive takes a syscall number, and some number of register-sized integer arguments,
and arranges to call that syscall in the target process,
returning the response as a single register-sized integer.

\texttt{rsyscall}'s default userspace cross-process syscall primitive sends the syscall request
to a userspace ``syscall server'' running in the target process,
which performs the syscall and sends the response back.
Communication to the syscall server happens over file descriptors,
typically a pair of pipes.
We use file descriptors for transport of data rather than shared memory
to support straightforward integration with event loops.

ptrace provides a way to perform arbitrary actions on other processes,
including system calls,
but is not suitable for us.
Among other issues, it has the substantial disadvantage of not permitting multiple ptracers.
A ptrace-based implementation would prevent using \texttt{strace} or \texttt{gdb} on \texttt{rsyscall}-controlled processes,
which is an unacceptable limitation for a general-purpose utility.

@@

The most simple way to access memory is for the local process to be in the same address space as the target process.
This is the case most of the time; we pass \verb|CLONE_VM| to \texttt{clone} by default in the Python wrapper.

Sometimes, the target process may be in another address space;
for example, if the target process is at a different privilege level,
it must be in a different address space.
There are a number of available techniques in that scenario;
we choose to copy memory over a pair of pipes,
again using file descriptors for the sake of easy event-loop integration.
\subsection{\texttt{clone}}\label{clone}
Now that we've established the basic implementation details of \texttt{rsyscall},
we'll consider specific issues related to process creation and initialization.

Besides \texttt{clone}, \texttt{vfork} and \texttt{fork} also create processes,
but they are not suitable for flexible direct-style process creation.
\texttt{vfork} suspends execution of the parent process
while waiting for the child process to exit or call \texttt{execve},
which is immediately unsuitable.
\texttt{fork} lacks many features which are restricted to \texttt{clone}.
For example, with \texttt{fork},
we could not create child processes which share the parent's address space,
which would complicate memory access.

The raw \texttt{clone} system call creates a new process
which immediately starts executing at the next instruction after the syscall instruction,
in parallel with the parent process,
with its registers in generally the same state as the parent process.

\texttt{clone} lets us change the stack for the new process.
We can use this to make the new process call an arbitrary function,
by storing its address on the new stack.
Further arguments can also be passed on the stack,
with the aid of a trampoline to match C calling conventions if necessary.

We use this to create processes running the syscall server.
After this,
the parent process can be begin to call system calls in the child process.
Most system calls work as normal;
the new child process can be modified freely through \texttt{chdir}, \texttt{dup2}, and other system calls.
From the system calls related to process creation,
only \texttt{execve} and \texttt{unshare} need substantial further attention.
\subsection{\texttt{exec}}\label{execve}
\texttt{execve} is unusual and requires careful design,
because when it is successful, it does not return.
We need a way to determine if \texttt{execve} is successful;
naively waiting for a response to the syscall request may leave us waiting forever.

There is a traditional technique used with \texttt{fork}
to detect a successful \texttt{execve} using a pipe,
which unfortunately won't work for us.
In this technique, the parent process creates a pipe before forking,
ensures both ends are marked \verb|O_CLOEXEC|,
performs the fork,
closes the write end of the pipe,
and \texttt{read}s the read end of the pipe.
The child process either successfully calls \texttt{execve} or exits;
either way, the last copy of the write end of the pipe will be closed,
which causes the \texttt{read} in the parent process to return EOF.
The parent can then check that the child process hasn't exited;
if the child hasn't exited, then it must have successfully called exec.

This trick works with \texttt{fork},
but it's not general enough to work with \texttt{clone}.
Child processes can be created with the \verb|CLONE_FILES| flag passed to \texttt{clone},
which causes the parent process and child process to share a single fd table;
we showed an example of this in section \ref{shared_fd_table}.
This means that when the parent process closes the write end of the pipe,
it will also be closed in the child process,
and the read end of the pipe will immediately return EOF,
before the child has called \texttt{execve} or exited.

Fortunately, there is an alternative solution, which does work with \verb|CLONE_FILES|.
\texttt{clone} has an argument, \texttt{ctid}, which specifies a memory address.
If the \verb|CLONE_CHILD_CLEARTID| \texttt{clone} flag is set,
then when the child exits or execs,
the kernel will set \texttt{ctid} to zero and then,
crucially, perform a futex wakeup on it.

Futexes are usually used for the implementation of userspace shared-memory synchronization constructs,
but the relevant detail for us here is that we can wait on an address
until a futex wakeup is performed on that address.
This means we can wait on \texttt{ctid} until the futex wakeup is performed,
and in this way get notified of the child process calling \texttt{execve} or exiting.
A process can only wait on one futex at a time;
to monitor multiple futexes from a single event loop,
we need to create a dedicated child process for each futex we want to wait on,
and have this child process exit when the futex has a wakeup.
We can then monitor these child processes from an event loop using standard techniques.\cite{signalfd}\cite{pidfd}

So, we pass \texttt{ctid} whenever we call \texttt{clone},
and set up a process to wait on that futex.
Then, when we call any syscall,
we wait for either the syscall to return an error or the futex process to exit,
whichever comes first.
Either will normally indicate that the syscall fails,
but if the futex process exits,
without the child process itself exiting,
then we can catch that in \texttt{execve} and know that the child has successfully called \texttt{execve}.

If the futex process and child process both exit,
it's ambiguous whether the child process successfully called \texttt{execve};
this ambiguity is unfortunate, but it is also present in the pipe-based approach.

Some other Unix-like systems natively provide the ability to wait for a child's \texttt{execve};
our implementation would be simplified, and the ambiguity eliminated,
if we had this ability on Linux.\cite{freebsd_kqueue}
One approach would be to add a new \texttt{clone} flag to
opt-in to receiving \texttt{WEXECED} events through \texttt{waitid}.
A new \texttt{waitid} flag alone is not sufficient,
since, to integrate this feature into an event loop,
it's necessary to receive \texttt{SIGCHLD} signals or readability notifications on a pidfd
when the \texttt{WEXECED} event happens.
\subsection{Managing file descriptor tables}\label{fdtables}
\texttt{clone} (without \verb|CLONE_FILES|) and \texttt{fork}
deal with file descriptor tables in the same way:
The new child process has a new file descriptor table,
containing copies of all the file descriptors existing in the parent at the time of the system call.
This is known as file descriptor inheritance.

As demonstrated in section \ref{shared_fd_table},
when the \verb|CLONE_FILES| flag is passed to \texttt{clone},
the parent process and the child process will share a single file descriptor table,
so changes to the file descriptor table made in one process will affect the other.

The file descriptor table can be unshared by calling \verb|unshare(CLONE_FILES)| or \texttt{execve}.
After either system call, the process will have a new file descriptor table
which is no longer shared with any other process.
All the file descriptors existing in the old table at the time of the system call
will be copied into the new file descriptor table,
except for, in the case of \texttt{execve}, file descriptors with the \texttt{CLOEXEC} flag set.
\subsubsection{Tracking file descriptors through handles}
After \texttt{clone}, \verb|unshare(CLONE_FILES)|, or \texttt{execve}
has copied all the file descriptors from some old table into a new table,
we need to decide which file descriptors we want to keep open in the new table,
and which file descriptors should be closed.

In general, if a certain file descriptor is in use by the program,
through a certain process,
that file descriptor should stay open in that process's file descriptor table.
